---
title: "Linear and Logistic Regression Cheat Sheet"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---


```{r, warning = FALSE, message = FALSE}
library(rstatix)
library(tidyverse)
library(pastecs)
library(knitr)
library(epitools)
library(jtools)
library(Epi)
library(gtsummary)
options(scipen=999) 
```

### Open the Data

We are interested in the association between gender and BMI. We will do descriptive analysis and examine associations. This is a cheat sheet so mostly to show code examples and is not specific to your assignments. 

```{r}
data <- read_csv("Data.csv")
```

### Recode and label categorical variables as necessary. 

We saw this in data work week 2. [https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week2_data_work_R.md](https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week2_data_work_R.md)

```{r}
data <- data %>%
	mutate(gender_recode = case_when(
		SDC_GENDER == 1 ~ "Male",
    SDC_GENDER == 2 ~ "Female"
	))

data$gender_recode <- as.factor(data$gender_recode) 

### Checking our code to make sure our recode is correct
table(data$SDC_GENDER, data$gender_recode)
```

Calculate the percents and frequencies 

```{r}
## First we need to make sure we get the variables recoded and organized

### Example with gender where we already recoded
gender_table <- bind_rows(table(data$gender_recode), prop.table(table(data$gender_recode)))
```

### Visualize the data

Nothing fancy here just default ggplot2. Inspect of the type of distribution, outliers, and anything else. Probably need to fix a few things with the BMI variable. 

We saw this in data work week 3. [https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week3_data_work_R.md](https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week3_data_work_R.md)

```{r}
bmi_histo <- ggplot(data, aes(PM_BMI_SR)) + 
              geom_histogram()
plot(bmi_histo)
```

Inspect distribution by gender to see if anything is different other what we would expect. 

```{r}
boxplot <- ggplot(data, aes(PM_BMI_SR)) + 
              geom_boxplot() + 
              coord_flip() + 
              facet_wrap(~ gender_recode)
plot(boxplot)
```

### Examine bi-variable associations 

We saw this in data work week 4 [https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week4_data_work_R.md](https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week4_data_work_R.md) and 5 [https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week5_data_work_R.md](https://github.com/walkabilly/chep801_usask/blob/main/Data%20Work/Week5_data_work_R.md). 

#### Linear regression

```{r}
bmi_gender <- glm(PM_BMI_SR ~ gender_recode, data = data, family = "gaussian")
summary(bmi_gender)
```

#### Logistic regression

```{r}
gender_bmi <- glm(gender_recode ~ PM_BMI_SR, data = data, family = "binomial")
summary(gender_bmi)

multi_table <- tbl_regression(gender_bmi, exponentiate = TRUE) 

multi_table %>% as_kable()
```

Check model fit and assumptions

We saw this in data work week 5 and the model assumptions lecture.

```{r}
plot(gender_bmi)
```


