---
title: "Week 2 Data Work"
output:
      html_document:
        keep_md: true
---


```{r, warning = FALSE, message = FALSE}
library(rstatix)
library(tidyverse)
library(pastecs)
library(knitr)
library(epitools)
library(Epi)
options(scipen=999) 
```


```{r}
data <- read_csv("CANPATH_data_wrangling.csv")
```

## General instructions

### 1. Identify the variable types for each variable in the dataset

Example

__R__

```{r}
glimpse(data)
```

All of the variables here currently coded as _dbl_ except `ID` which is _chr_. That's incorrect and we are going to need to fix that when we recode.

### 2. Recode and label categorical variables as necessary

__R__

```{r}
data <- data %>%
	mutate(gender_recode = case_when(
		SDC_GENDER == 1 ~ "Male",
    SDC_GENDER == 2 ~ "Female"
	))

data$gender_recode <- as.factor(data$gender_recode) 

### Checking our code to make sure our recode is correct
table(data$SDC_GENDER, data$gender_recode)
```

### 3. Add the `NUT_VEG_QTY` and the `NUT_FRUITS_QTY` to create one variable
    * Create a categorical variable that represents the recommended fruit and vegetable consumption per day

__R__

```{r}
summary(data$NUT_VEG_QTY)
summary(data$NUT_FRUITS_QTY)

### No missing data the no weird codings for the numbers. Great! 

data <- data %>%
	mutate(fruit_veg_tot = NUT_VEG_QTY + NUT_FRUITS_QTY)

summary(data$fruit_veg_tot)

data <- data %>%
	mutate(fruit_veg_cat = case_when(
		fruit_veg_tot <= 7  ~ "Not Meeting Guidelines",
    fruit_veg_tot > 7 ~ "Meeting Guidelines"
	))

table(data$fruit_veg_cat)

data <- data %>%
	mutate(fruit_veg_dic = case_when(
		fruit_veg_tot <= 7 ~ 0,
    fruit_veg_tot > 7 ~ 1
	))

table(data$fruit_veg_tot, data$fruit_veg_cat)
```

### 4. Create a categorical variable for the `PA_TOTAL_SHORT` variable the defines low, moderate, and high activity per week
    * Low - < 600 MET minutes per week
    * Moderate - >=600 to <3000 MET minutes per week
    * High - >=3000 MET minutes per week

### 5. Calculate the mean and standard deviation
    * For the variables were it is appropriate
    * Including the new variables you have created
    
```{r}
summary(data$PA_TOTAL_SHORT)

data <- data %>%
	mutate(pa_cat = case_when(
		PA_TOTAL_SHORT < 600  ~ "Low Activity",
		PA_TOTAL_SHORT >= 3000 ~ "High Activity",
		PA_TOTAL_SHORT >= 600 ~ "Moderate Activity"
	))

### Flag we need to put the low and how activity first otherwise we don't get the right answer
```

### 6. Calculate the percents and frequencies 
    * For the variables were it is appropriate to do so

There are __MANY__ different ways to do this in R. There is no one best way. 

```{r}
## First we need to make sure we get the variables recoded and organized

### Example with gender where we already recoded
gender_table <- bind_rows(table(data$gender_recode), prop.table(table(data$gender_recode)))

### Example with fruit_veg_cat where we already recoded
fruit_veg_table <- bind_rows(table(data$fruit_veg_cat), prop.table(table(data$fruit_veg_cat)))
```

### 7. Are there missing data?

```{r}
summary(data$fruit_veg_tot)
```

### 8. Create a 2*2 table for fruit and vegetable consumption by gender
    * Interpret the 2*2 table result

```{r}
gender_fv_table <- table(data$gender_recode, data$fruit_veg_cat)
gender_fv_table

epitab(gender_fv_table, method = "oddsratio")
```

## OOOPSSSS 

#### Epitab expects

| --- | --- | Disease | --- | 
| --- | --- | --- | --- | --- |
| --- | --- | No (REF) | Yes |
| Exposure | Level 1 (REF) | a | b | 
| --- | Level 2 | c | d | 

#### We gave it 

| --- | --- | Disease | --- | 
| --- | --- | --- | --- | --- |
| --- | --- | Yes  | No (REF) |
| Exposure | Level 1 (REF) | b | a | 
| --- | Level 2 | d | c | 

```{r}
gender_fv_table <- table(data$gender_recode, data$fruit_veg_cat)
gender_fv_table

epitab(gender_fv_table, method = "oddsratio", rev = "columns") ## Here we flip the columns around to get the right answer for males. 
```
   