---
title: "Bias Assignment"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load the episenr package after installation. This needs to be done every time you analyse
library("episensr") #you can quote or unquote, but need quote when using install.package, but not when using library, don't ask me why
library(htmlTable)

# packaged used in the data work lab and assignment 1, needed to run some functions in this assignment
library(tidyverse)
library("knitr")
library("epitools")
```


First, lets load the data. Make sure to specify your own path to the data. 
```{r, warning = FALSE}
data <- read_csv("/Users/dlf545/Dropbox/Teaching/USask/CHEP 801/chep801_usask/Data.csv")
```


# Question 1a
#### Enter the selection probabilites provided below into the `bias_params` line in the episens command  and show the output. You need to copy and paste the `selection` function of episens I wrote above, paste below, and enter the appropriate selection probabilities. THere is no need to edit anything else in the command.  

  - Exposed (low PA) and Outcome (Diabetes) = 0.9
  - Exposed and no Outcome =0.6
  - Unexposed and Outcome = 0.8
  - Unexposed and no Outcome = 0.8
```{r}
#Enter your answer here 
```


# Question 1b 
#### Based on the output you generated, is there selection bias in the original RR?
```{r}
#Enter your answer here 
```

# Question 1c 
#### Another study on the same population found out that the selection probabitlies are actually as follows: 
  - Exposed (low PA) and Outcome (Diabetes) = 0.9
  - Exposed and no Outcome =0.9
  - Unexposed and Outcome = 0.8
  - Unexposed and no Outcome = 0.8
```{r}
#Enter your answer here 
```

#### Run the `selection` command again (paste below) based on these updated probabilities by filling  `bias_parms`. Again, careful with the ordering of these values. 


          
# Question 1d
Based on the output from Question 1c, answer if there is selection bias or not, and if not please explain why in one or two sentences.
```{r}
#Enter your answer here 
```


# Question 2 - Information bias 
#### We will correct missclassificaiton of the exposure, PA variable. 
Suppose that a subsequent validation study using these data show that people inaccurately reports physical activity status, with the following sensitivity and specificity of exposure classification among diabetes cases and non-cases.  

```{r}
# Lines below just creates example tables 
contingencyTable <- data.frame(Outcome_Yes = c("a", "c"), Outcome_No = c("b", "d"))
rownames(contingencyTable) <- c("Exposure_yes", "Exposure_No")

contingencyTable %>% 
  addHtmlTableStyle(css.cell = c("width: 140;","width: 140;")) %>% 
  htmlTable(caption = "Table cell labels") 

```
  - Se_outcome: Sensitivity of exposure classification among outcome (i.e., extent of misclassification from a to c among outcome)  = 0.95 
  - Sp_outcome: Specificity of exposure classification among outcome (i.e., extent of misclassification from c to a among outcome)   =0.8 
  - Se_noOutcome: Sensitivity of exposure classification among no outcome (i.e., extent of misclassification from b to d among no outcome) = 0.95   
  - Sp_noOutcome: Specificity of exposure classification among no outcome (i.e., extent of misclassification from  d to b among no outcome) = 0.8   
  
  
As a practice, we will plug in these classification metrics into the `misclassification` function in `episenser`, when there is no missclassification (i.e., all Se and Sp values are 1.0). The probabilities values entering the `bias_parms` line is *Se_outcome, Se_noOutcome, Sp_outcome, Sp_noOutcome*.   
```{r}
misclassification(matrix(c(434, 1094, 4937, 16263), 
                 dimnames = list(c("Diabetes", "No Diabetes"), 
                                 c("Low PA", "moderate/high PA")), 
                 nrow = 2, byrow = TRUE), 
                 type = "exposure",
          bias_parms  = c(1.0, 1.0, 1.0, 1.0))
```

# Question 2a
#### Now, perform bias sensitivity analysis based on the imperfect classification accuracies provided above (values of sensitivities and specificities), by updating the four values in the `bias_parms` line. Show the output below. 
```{r}
#Enter your answer here 
```


# Question 2b
#### Comparing `Misclassification Bias Corrected Relative Risk` and `Observed Relative Risk` (the latter is the crude - uncorrected association). Is there information bias? 
```{r}
#Enter your answer here 
```

# Question 2c
#### Is this differential or non-differential missclassification? Please provide 1 or 2 sentence of explanation. 
```{r}
#Enter your answer here 
```


# Question 2d
#### 

  - Sensitivity of exposure classification among outcome (i.e., extent of misclassification from a to c among outcome)  = 0.9
  - Specificity of exposure classification among outcome (i.e., extent of misclassification from c to a among outcome)   =0.9 
  - Sensitivity of exposure classification among no outcome (i.e., extent of misclassification from b to d among no outcome) = 0.5   
  - Specificity of exposure classification among no outcome (i.e., extent of misclassification from  d to b among no outcome) = 0.9   


```{r}
#Enter your answer here 
```

# Question 2e
#### Is this differential or non-differential missclassification (results of Question 2d)? Please provide 1 or 2 sentence of explanation. 
```{r}
#Enter your answer here 
```
