---
title: "Regression_OLS_Assumption_FEVdata"
output:
  html_document:
    keep_md: yes
---

## Setup 
Load library, set figure size
```{r}
rm(list=ls()) # clear all workspace, very important to do this to refresh 
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=4.5, fig.height=4.5) 
library(tidyverse)
library(epitools)
library(htmlTable)
library(autoReg)
library(labelled)
```

### Data size for simulation and custom function to format table names 
```{r setup variables}
N = 200

funcColNames <- function(d){
  colnames(d)[colnames(d) == "term"] <- "Coefficient"
  colnames(d)[colnames(d) == "conf.low"] <- "ConfInt.Lower"
  colnames(d)[colnames(d) == "conf.high"] <- "ConfInt.Higher"
  return(d)
}
# Make shorter variables names 
funcColNames2 <- function(d){
  colnames(d)[colnames(d) == "term"] <- "Coef"
  colnames(d)[colnames(d) == "conf.low"] <- "CI.Lower"
  colnames(d)[colnames(d) == "conf.high"] <- "CI.Higher"
  d <- d %>% select(-c(p.value, statistic))
  return(d)
}
```


## Real data to check the assumptions 
FEV (Forced Expiration Volume) study, can be download from many sources    

* Some anlaysis examples [https://www.tandfonline.com/doi/full/10.1080/10691898.2005.11910559](https://www.tandfonline.com/doi/full/10.1080/10691898.2005.11910559)  
* An Exhalent Problem for Teaching Statistics, Kahn (2017) [https://doi.org/10.1080/10691898.2005.11910559](https://doi.org/10.1080/10691898.2005.11910559)

```{r}
url <- "http://www.emersonstatistics.com/datasets/fev.txt"
download.file(url, "fev.txt" )
dat <- read.csv("https://github.com/walkabilly/chep801_usask/raw/main/fev.txt", sep="")

# Describe data 
glimpse(dat)

# Make categorical varaibles out of characters 
dat$sex <- factor(dat$sex)
levels(dat$sex) = c("Female", "Male")
table(dat$sex)

dat$smoke <- factor(dat$smoke)
levels(dat$smoke) <- c("Smoker", "Non-smoker")
table(dat$smoke)

pairs(dat %>%  select(-c("seqnbr", subjid)))
```


We will start from descriptive statistics
### We will assess the association of each variables as well as potential non-linearity and effect measure modification 
Descritpive plots - histogram 
```{r fig1, fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
# Distribution of continuous outcomes 
hist(dat$age, main = "Age", cex.main = 1)
hist(dat$fev, main = "FEV, outcome", cex.main = 1)
hist(dat$height, main = "Height", cex.main = 1)
```

### Descriptive plots - xy age and fev
```{r fig2, fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
ggplot(dat, aes(x= age, y = fev)) + 
  geom_point() +
  theme_classic() + 
  ggtitle("Relationship between Age (years) and FEV (Litres)") + 
  theme(plot.title = element_text(size = 9)) + 
   geom_smooth(se=FALSE)
  
ggplot(dat, aes(x= age, y = fev)) + 
  geom_point() +
  theme_classic() + 
  ggtitle("Relationship between Age (years) and FEV (Litres) \n  by Sex") + 
  facet_wrap(~sex) + 
  theme(plot.title = element_text(size = 9)) + 
   geom_smooth(se=FALSE)
```

### Descriptive plots - height and fev 
```{r fig3, fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
ggplot(dat, aes(x= height, y = fev)) + 
  geom_point() +
  theme_classic() + 
  ggtitle("Relationship between Height (years) and FEV (Litres)")+ 
  theme(plot.title = element_text(size = 9)) + 
   geom_smooth(se=FALSE)

ggplot(dat, aes(x= height, y = fev)) + 
  geom_point() +
  theme_classic() + 
  ggtitle("Relationship between Height (years) and FEV (Litres) \n by smoking status") + 
  facet_wrap(~smoke)+ 
  theme(plot.title = element_text(size = 9)) + 
   geom_smooth(se=FALSE)

ggplot(dat, aes(x= age, y = height)) + 
  geom_point() +
  theme_classic() + 
  ggtitle("Relationship between Age (years) and FEV (Litres)")+ 
  theme(plot.title = element_text(size = 9)) + 
   geom_smooth(se=FALSE)
```

### Descriptive plots - height and fev 
```{r fig4, fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
ggplot(dat, aes(sex, fev)) + geom_boxplot() + theme_classic() + ggtitle("FEV by sex")
ggplot(dat, aes(smoke, fev)) + geom_boxplot() + theme_classic() + ggtitle("FEV by smoking status")

knitr::kable(table(dat$sex), digits = 2, caption = "Results: Linear model ")
knitr::kable(table(dat$smoke), digits = 2, caption = "Results: Linear model ")

```

### Plot all varaibles at once, execpt person IDs 
```{r}
dat %>% select(-c(seqnbr, subjid)) %>% pairs
```


## Exploratory regression analysis,  
#### single variable regression between FEV and each covariate
#### remember though, just looking at coefficients only is not as imformative as x-y plots above 
```{r}
independentVariables <- c("age", "height", "smoke", "sex")
independentVariablesData<- dat[, independentVariables]
fitList <- lapply(1:ncol(independentVariablesData), function(x) lm(dat$fev ~ independentVariablesData[,x]))

# extract just coefficients
sapply(fitList, coef)

fit=lm(fev~age+sex+height+smoke,data=dat)

#fit=lm(fev~age*sex*height*smoke,data=dat)
autoReg(fit,uni=TRUE, threshold=0.2) %>% myft()

final=step(fit,trace=0)
summary(final)

```





#### Multivariable regression
Our model of interest is: 
$$E[Y_{fev}] = \alpha + \beta_{age}*age + \beta_{height}*height + \beta_{smoking}*smoking +  \beta_{sex}*sex  $$
```{r}
fit=lm(fev~age+sex+height+smoke,data=dat)
datPlot <- data.frame(dat, fitted =  fit$fitted.values, residual = fit$residuals)
```


### Our FEV model, for now throw everything, Exposure of interesting smoking yes/no 
#### As usual, be sure to consult with DAF for real analysis 
#### No interaction term for now 

### Assumption 1 - Lienarity of outcome with fitted values (or x if univariate model)
```{r}
ggplot(data = datPlot, aes(y = fev, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs Observed outcome") + 
  geom_smooth(se = FALSE) + 
  geom_abline(slope = 1,linetype = "dashed")


ggplot(data = datPlot, aes(y = residual, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs residuals") + 
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0, linetype = "dashed")


```

### Assumption 1 - plot by Sex and Smoking
#### Besure though, smokign is our exposure of interest, so we cannot subset data by smoke 
```{r}
ggplot(data = datPlot, aes(y = fev, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  facet_wrap(~smoke) + 
  ggtitle("Assumption 1: Fitted vs Observed outcome, \n by Smoke")+ 
  geom_abline(slope = 1,linetype = "dashed")


ggplot(data = datPlot, aes(y = fev, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  facet_wrap(~sex) + 
  ggtitle("Assumption 1: Fitted vs Observed outcome, \n by Sex")+ 
  geom_abline(slope = 1,linetype = "dashed")

```

### Assumption 1 - Perhaps, quadratic term in the regression model, since we saw that FEV is non-linear with age and height? This cold lead to residual confounding by age and height. 
#### Abter adding quadratic term for age and height, the association of fitted and observed FEV look linear **BUT**, the interpretation of age and height are no longer linear (quadratic term added), and there is a larger disagreement at higher values of fitted and observed. 
```{r assumption 1 fig quadratic , fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
dat$age2 <- dat$age*dat$age
dat$height2 <- dat$height*dat$height
fit=lm(fev~age +age2 +sex+height + height2+smoke,data=dat)
datPlot_Q <- data.frame(dat, fitted =  fit$fitted.values, residual = fit$residuals)

ggplot(data = datPlot_Q, aes(y = fev, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs Observed outcome, \n quadratic age and height")


ggplot(data = datPlot_Q, aes(y = fev, x = fitted, color = smoke)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs Observed outcome, \n quadratic age and height \n color separated by smoking ")

ggplot(data = datPlot, aes(y = fev, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs Observed outcome \ Original model ")

```



### Assumption 2 - Independence of errors (residuals)
```{r}

```


#### Assumption 3 - Constant dispersion of residuals across the value of fitted values (if multiple regression), or x (if univariate model) 
```{r assumption 3 , fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
ggplot(data = datPlot, aes(y = residual, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 3: Fitted vs residuals \n increase of variance \n but U-shaped") + 
  geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0, linetype = "dashed")

ggplot(data = datPlot_Q, aes(y = residual, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 3: Fitted vs age \n increase of variance") + 
    geom_smooth(se = FALSE) + 
  geom_hline(yintercept = 0, linetype = "dashed")
```

#### Assumption 3 - Constant dispersion of residuals across the value of fitted values (if multiple regression), or x (if univariate model) 
```{r assumption 3 by smoking subset, fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
ggplot(data = datPlot[datPlot$smoke == "Smoker", ], aes(y = residual, x = fitted)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 3: Fitted vs height \n smokers only") + 
  geom_hline(yintercept = 0, linetype = "dashed")

```


#### Assumption 3 - Constant dispersion of residuals across the value of fitted values (if multiple regression), or x (if univariate model) 
```{r assumption 3 by covarites, fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
ggplot(data = datPlot, aes(y = residual, x = age)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs age \n increase of variance") + 
  geom_hline(yintercept = 0, linetype = "dashed")

ggplot(data = datPlot, aes(y = residual, x = height)) + 
  geom_point() + 
  theme_classic() + 
  ggtitle("Assumption 1: Fitted vs height \n increase of variance \n and bit of u-shape") + 
  geom_hline(yintercept = 0, linetype = "dashed")
```


### Assumption 4 - Resisuals are normally distributed 
#### We typically do not get very normal distribution of residuals 
```{r assumption 4 , fig.height = 4, fig.width = 4, fig.show="hold", out.width="50%"}
hist(fit$residuals, main = "Histrogram of residuals, longer tails", breaks = 20)
qqnorm(fit$residuals)
qqline(fit$residuals)
```

### Assumptions - Summary of residual diagnosis 
```{r}
plot(fit)
```









